Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2017-06-21 2017-Xudong-SMT/submit-ICTAC2017.tex
reduce section 2

Implementing this semantics raised several challenges, in order:
\begin{itemize}
  \item To get a tool that could be applied to pNets representing
    various languages, in particular various actions algebras,
    with their specific decision theories.
  \item To separate clearly the algorithmic part, as a residual
      algorithm generating the transitions of the open automaton from
      combination of all possible (symbolic) behaviors on one hand; from the
      symbolic reasonning part, specificaly here checking the
      satisfiability of the predicates generated by our algorithm,
      that we delegate to a generic SMT engine, on the other hand.
  \item To build a prototype and validate the approach on our basic
    case-studies, and understand the efficiency of the interaction
    with the SMT solver.
\end{itemize}

@@

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item We have defined the open automaton generation algorithm, and we
  implemented a full working prototype, within the 
  VerCors platform. In the process, we had to improve significantly the
  semantics rules from~\cite{henrio:Forte2016}, and add features in
  the algorithm to deal the full 
  model, including management of variables and assignments.
%  \item We extended the meta-model of action expressions, and the
%    associated API, in the VerCors platform, to be able to manage
%    parameterisation over the action algebra operators.
\item We have implemented the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the logical predicates
    generated by the algorithm.
\item We validated the whole approach on small use-cases covering all
      important features of the model.
\end{itemize}

@@

In this paper, we use an example inspired from LOTOS, to illustrate
the main features of pNets and the way we deal with them in the symbolic
semantics, and in the satisfiability checking. 
%% \QIN{
%% Meanwhile, 
%% the formula using operators from CCS can also be encoded into a pNet and 
%% conducted with the same things.
%% }
%% \subsection{LOTOS Formula Example}
We choose here the expression \texttt{P>>(acc(v);Q)}, that includes the
Enable and Prefix 
operators of the LOTOS language. For the readers not familiar with
Lotos, the pNet encodings here should be sufficient to understand the
Enable operator semantics.
In Fig.~\ref{schema:lotos-pnet}, we show the encoding of the formula using pNets. 
This pNet is composed of: a toplevel node PN1 encoding the Enable
operator, with its controller (pLTS) C1; a hole representing the
process variable P; another node PN2 representing the Prefix
operator, with its controller C2, and its hole Q.

In order to show all features of the pNet model, we choose to use two
styles of encodings in the controllers.
The most common style used in process calculi is ``state oriented'',
we use it here for C2: it has two states to
indicate the position of the control point in the
Prefix expression $(acc(v);Q)$. The second one is
``data oriented'': controller C1 has a single state, with a state
variable $s_0$ encoding the change of control point from the left to
the right subnet.
%% The transitions in the pLTS have the more general syntax:
%% ``<action-expr> [guard] {sequence of assignments}''. Guards and
%% assignments can only use the values of variables of their source state
%% and values from input variables in the action expression; and a
%% transition can only assign variables of its target state.

Now consider the top pNet node PN1. In addition to
its controller it has 2 sub-pNets: the hole P and the sub-pNet PN2.
Three synchronization vectors define the possible
synchronizations between the actions of its subnets.
The first vector ``$< l, a_1 , \_ > -> a_1$'', transmitting an action
$a_1$ of P, can only occur if $a_1$ is not  
a $\delta$, here ``$\_$'' means that the corresponding subnet is
not involved in this synchronization, and the ``l'' action is related
to the corresponding transition of the controller C1: this action is
available only when the state variable $s_0$ is equal to 0.
The second synchronisation vector composes an
action $\delta(x)$ from $P$ with an $acc(x)$ from $Q$, transmitting a
value $x$ before $Q$ takes the control; 
The vector result $\underline{\delta(x)}$ is an
internal action, called ``synchronized'' that cannot be further
synchronized at upper levels. Synchronized actions is a straightforward
generalization of the notion of internal actions, that will be
convenient for observing internal events during model-checking.
The last vector acts is to the first one.  
Similarly the node PN2 has two synchronization vectors to encode
the behavior of the Prefix operator.

@@

This definition is not constructive but it is easy to build the
predicate constructively by brute-force unification of the sub-pNets
actions with the corresponding vector actions, possibly followed by a simplification
step.

@@

\paragraph{Structural Semantic Rules:}
Now we build the semantics of an open pNet as an open automaton where
LTSs are the pLTSs at 
the leaves of the pNet structure, and the states are structured as in
the previous section. 
%given by Definition~\ref{def-states}.
The open transitions first
 project the global state into states of the leaves, then apply
pLTS transitions on these states, and compose them with the sort of the holes. %The pNet
%structure does not appear in the open-automaton, only the
%set of Holes and the set of Leaves.
The semantics   regularly instantiates \emph{fresh} variables, and uses a
\emph{clone} operator that clones a term replacing each variable with a
fresh one.
The variables in each synchronization vector are considered local:
for a given pNet expression, we must have fresh local variables for
each occurrence of a vector (= each time we instantiate rule
Tr2). Similarly the state variables of each copy of a
given pLTS in the system, must be distinct, and those created for each
application of Tr2 have to be fresh and all distinct.

@@

  The rule for a pLTS $p$  checks that the guard 
  is verified and transforms assignments into post-conditions:

  \begin{description}
    \item[{\bf Tr1:}]
    $\inferrule
    { s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
          J}\rangle} s'\in \to  \\
      {\tt fresh}(v) \\
      Pred = ~e_b \land (\alpha = v)
    }
    { p = \mylangle  S,s_0, \to \myrangle
      \models
      \inferrule*[myfraction=\reddottedrule]
      {\{s \xrightarrow{\alpha}_p s'\} ,\emptyset ,
      Pred,\left\{x_j\gets e_j\right\}^{j\in J}}
      {\ostate{s} \xrightarrow{v} \ostate{s'}}
    }
    $
  \end{description}
% Note that this note is greatly simplified by the fact that variables are local to 
% thread; introducing global state variables or accepting loops to the same 
% state would 
% require to reason 
% on the scope of 
% each variables, and to introduce additional variables to handle the several occurence 
% of the same pLTS variable in the predicates. Indeed the constraints on pLTS 
% transitions 
% ensure that the same variable never appears both on the left and on the right of the 
% equations of a predicate.

@@

\State Initialize a set for unexplored global states $U$, a set for explored global states $E$;
\While{!isEmpty(U)}
	\State Extract a global state $S$ from the $U$;
	\State Start at the root pNet node
	\State Recursively applying Tr1 or Tr2 on it and its subnets;
	\If{Subnet is pLTS, apply the Tr1} 
		\For{each transition $TR$ start from the state $s$ in $S$}
			\State Generate open transition $OT$ according to $TR$;
			\State Store $OT$ into the result list $L$;
		\EndFor
	\EndIf
	\If{Subnet is pNet, apply the Tr2} 
		\For{each $sv \in SV$}
			\State Invoke $Combining()$ to generate the combination of sub-transitions;
			\State Invoke $Matching()$ to match the combination with the $sv$;
			\State Store the result set of open transitions into $L$;
		\EndFor
	\EndIf
	\For{each target global state $T$ from $OT \in L$}
		\IfThen{$(!U.contains(T))\ \&\&\ (!E.contains(T))$}{Add $T$ into the $U$;}
	\EndFor
\EndWhile
\State Prune the open transitions in $L$ using SMT solver;
\State \Return $L$;

@@

\subsection{Combining and Matching using TR2}
\TODO{Xudong: move here a (new version) of the alogorithm loplevel sketch}
\QIN{[Move to the head of Section 5.]}

>>>>

Implementing this semantics raised several challenges, in order:
\begin{itemize}
  \item to get a tool that could be applied to pNets representing
    various languages, in particular various actions algebras,
    with their specific decision theories,
  \item to separate clearly on one hand the algorithmic part, as an
      algorithm generating the transitions of the open automaton from
      combination of all possible (symbolic) behaviors; on the other hand
      the symbolic reasonning part, specificaly here using an SMT
      engine to check the
      satisfiability of the predicates generated by our algorithm,
  \item to build a prototype and validate the approach on our basic
    case-studies, and understand the efficiency of the interaction
    with the SMT solver.
\end{itemize}

@@

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item We define the open automaton generation algorithm, and we
  implemented a full working prototype, within the 
  VerCors platform. In the process, we improved the
  semantics rules from~\cite{henrio:Forte2016}, and add features in
  the algorithm to deal the full 
  model, including management of variables and assignments.
%  \item We extended the meta-model of action expressions, and the
%    associated API, in the VerCors platform, to be able to manage
%    parameterisation over the action algebra operators.
\item We implement the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the transitions
    generated by the algorithm.
\item We validate the whole approach on small use-cases covering all
      important features of the model.
\end{itemize}

@@

In this paper, we use an example inspired from LOTOS, to illustrate
the main features of pNets and the way we deal with them in the symbolic
semantics, and in the satisfiability checking. 
%% \QIN{
%% Meanwhile, 
%% the formula using operators from CCS can also be encoded into a pNet and 
%% conducted with the same things.
%% }
%% \subsection{LOTOS Formula Example}
We choose here the expression \texttt{P>>(acc(v);Q)}, that includes the
Enable and Prefix 
operators of the LOTOS language. For the readers not familiar with
Lotos, the pNet encodings here should be sufficient to understand the
Enable operator semantics.
In Fig.~\ref{schema:lotos-pnet}, we show the encoding of the formula using pNets. 
This pNet is composed of: a toplevel node PN1 encoding the Enable
operator, a pLTS C1 acting as the {\sl controller} of the node; a hole
representing the 
process variable P; another node PN2 representing the Prefix
operator, with its controller C2, and its hole Q.

In order to show all features of the pNet model, we choose to use two
styles of encodings in the controllers.
% The most common style used in process calculi is ``state oriented'',
Controller C2 is ``state oriented'': it has two states to
indicate the position of the control point in the
Prefix expression $(acc(v);Q)$. The second one is
``data oriented'': controller C1 has a single state, with a state
variable $s_0$ encoding the change of control point from the left to
the right subnet.
%% The transitions in the pLTS have the more general syntax:
%% ``<action-expr> [guard] {sequence of assignments}''. Guards and
%% assignments can only use the values of variables of their source state
%% and values from input variables in the action expression; and a
%% transition can only assign variables of its target state.

The root pNet node PN1 has 3 sub-pNets: its controller C1,
the hole P and the sub-pNet PN2.
Three synchronization vectors define the possible
synchronizations between the actions of the subnets.
The first vector ``$< l, a_1 , \_ > -> a_1$'' transmits an action
$a_1$ of P to the upper level. Here the ``l'' action is related
to the corresponding transition of the controller C1, ``$\_$'' means
that the corresponding subnet is not involved in this synchronization,
and there is a guard meaning this synchronisation can only occur if
$a_1$ is not a $\delta$.
The second synchronisation vector composes an
action $\delta(x)$ from $P$ with an $acc(x)$ from $Q$, transmitting a
value $x$ before $Q$ takes the control; 
The vector result $\underline{\delta(x)}$ is a ``synchronized''
action, that cannot be further
synchronized at upper levels. Synchronized actions is a useful
generalization of the notion of internal actions, that will be
convenient for observing internal events during model-checking.
% The last vector acts is to the first one.  
% Similarly the node PN2 has two synchronization vectors to encode
% the behavior of the Prefix operator.

@@

%% This definition is not constructive but it is easy to build the
%% predicate constructively by brute-force unification of the sub-pNets
%% actions with the corresponding vector actions, possibly followed by a simplification
%% step.

@@

\paragraph{Structural Semantic Rules:}
Now we build the semantics of an open pNet as an open automaton where
LTSs are the pLTSs at 
the pNet leaves, and the states are structured as in
the previous section. 
%given by Definition~\ref{def-states}.
To build an open transition one first
 projects the global state into states of the leaves, then applies
pLTS transitions on these states, and compose them with actions of
holes using synchronisation vectors. %The pNet
%structure does not appear in the open-automaton, only the
%set of Holes and the set of Leaves.

The semantics   regularly instantiates \emph{fresh} variables, and uses a
\emph{clone} operator that clones a term replacing each variable with a
fresh one.
The variables in each synchronization vector are considered local:
for a given pNet expression, we must have fresh local variables for
each occurrence of a vector (= each time we instantiate rule
Tr2). Similarly the state variables of each copy of a
given pLTS in the system, must be distinct, and those created for each
application of Tr2 have to be fresh and all distinct.

@@

  The rule ({\bf Tr1}) for a pLTS $p$  checks that the guard 
  is verified and transforms assignments into post-conditions:

  \begin{description}
    \item[{\bf Tr1:}]
    $\inferrule
    { s \xrightarrow{\langle \alpha,~e_b,~(x_j\!:= {e}_j)^{j\in
          J}\rangle} s'\in \to  \\
      {\tt fresh}(v) \\
      Pred = ~e_b \land (v = \alpha)
    }
    { p = \mylangle  S,s_0, \to \myrangle
      \models
      \inferrule*[myfraction=\reddottedrule]
      {\{s \xrightarrow{\alpha}_p s'\} ,\emptyset ,
      Pred,\left\{x_j\gets e_j\right\}^{j\in J}}
      {\ostate{s} \xrightarrow{v} \ostate{s'}}
    }
    $
  \end{description}
% Note that this note is greatly simplified by the fact that variables are local to 
% thread; introducing global state variables or accepting loops to the same 
% state would 
% require to reason 
% on the scope of 
% each variables, and to introduce additional variables to handle the several occurence 
% of the same pLTS variable in the predicates. Indeed the constraints on pLTS 
% transitions 
% ensure that the same variable never appears both on the left and on the right of the 
% equations of a predicate.

@@

\State Initialize sets $U$, $E$ for unexplored/explores global states
$U$, $L$ for result OTs;
\While{!isEmpty(U)}
	\State Extract a global state $S$ from the $U$;
	\State Recursively applying Tr1 or Tr2 on the root node and its subnets:
	\If{Subnet is pLTS}
                \State $s$ is the local state of the pLTS in the global state $S$;
		\For{each transition $tr$ starting from $s$}
			\State apply Tr1 to build the corresponding
                        $OT$; store $OT$ in $L$;
		\EndFor
	\EndIf
	\If{Subnet is pNet, apply the Tr2} 
		\For{each $sv \in SV$}
			\State Invoke $Combining()$ to generate the combination of sub-transitions;
			\State Invoke $Matching()$ to match the combination with the $sv$;
			\State Store the result set of open transitions into $L$;
		\EndFor
	\EndIf
	\For{each target global state $T$ from $OT \in L$}
		\IfThen{$(!U.contains(T))\ \&\&\ (!E.contains(T))$}{Add $T$ into the $U$;}
	\EndFor
\EndWhile
\State Prune the open transitions in $L$ using SMT solver;
\State \Return $L$;

@@

\subsection{Combining and Matching using TR2}
% \TODO{Xudong: move here a (new version) of the alogorithm loplevel sketch}
% \QIN{[Move to the head of Section 5.]}
